---
title: "Addendum: Illustrative predictive models"
author: "John Lövrot"
license: "CC BY 4.0"
date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 5
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = 'graphics/addendum-illustrative-models-',
  echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load_project}
setwd("..")
ProjectTemplate::reload.project()
setwd("reports")

library(caret)
library(glmnet)
library(randomForest)
library(pROC)
```

```{r}
ggplot2::theme_set(theme_classic() +
    theme(axis.line.x = element_blank()) + 
    theme(axis.line.y = element_blank()))
trellis.par.set(caretTheme())
options(knitr.kable.NA = "")
```

```{r}
pData(casecontstudy) <- pData(casecontstudy) %>%
  as_tibble() %>%
  mutate(
    h2stat = factor(subtypecd %in% "H2", 
      levels = c(FALSE, TRUE), labels = c("other", "H2"))
  ) %>%
  (function(x) data.frame(x, row.names = x$subjid))

stopifnot(validObject(casecontstudy))
```

```{r derive_HER2_amplicon_metagene}
sel_symbols <- c("ERBB2", "GRB7")

eset <- casecontstudy[fData(casecontstudy)$symbol %in% sel_symbols, ] %>%
  genefilter::featureFilter()
featureNames(eset) <- as.character(fData(eset)$symbol)
pData(eset) <- cbind(pData(eset), t(exprs(eset)))
casecontstudy$HER2_amplicon_metagene <- colMeans(exprs(eset))

eset <- qcsubstudy[fData(qcsubstudy)$symbol %in% sel_symbols, ] %>%
  averageExprsByFDataVar("symbol")
featureNames(eset) <- as.character(fData(eset)$symbol)
qcsubstudy$HER2_amplicon_metagene <- colMeans(exprs(eset))
```

```{r}
load(file.path("..", "cache", "caret", "training.RData"))
```

# Illustrative predictive models

## Model training

The full study is split into a training set (2/3) and a test set (1/3), 
with the QC substudy as part of the test set.

```{r}
casecontsets <- unique(casecontstudy$setnr)
tibble(
  casecontset = casecontsets, 
  partition = factor(
    casecontset %in% pData(casecontstudy)[rownames(training), "setnr"], 
    levels = c(TRUE, FALSE), 
    labels = c("Training set", "Test set")), 
  in_qcsubstudy = factor(
    casecontset %in% qcsubstudy_primary_endpoint_inclusion$setnr, 
    levels = c(TRUE, FALSE), 
    labels = c("QC substudy", "(rest)"))) %>%
  dplyr::select(in_qcsubstudy, partition) %>%
  table() %>%
  addmargins() %>%
  knitr::kable(caption = "Table. Number of case-control sets in each partition of the full study.")
```

A panel of models are trained and tuned based on the training set using repeated cross-validation, a resampling technique.
For details, see [reproduce-cunha15canres/addendum-train-illustrative-models.html](./addendum-train-illustrative-models.html)

## Performance in training and test sets

```{r}
resamps <- resamples(fit_list)
```

```{r}
## Select model based on comparison of performance in training set
if (interactive()) 
  dotplot(resamps, metric = "ROC")

selected_modelnam <- "Logistic Regression with Elastic Net Regularisation"
fit <- fit_list[[selected_modelnam]]
```

Selected model: 
```{r, results = "asis"}
cat(selected_modelnam)
```

```{r}
resamps_tbl <- resamps %>%
  as.data.frame() %>%
  tidyr::gather(varnam, value, -Resample) %>%
  tidyr::separate(Resample, c("fold", "rep"))

## Find ordering of models
tmp <- resamps_tbl %>% 
  group_by(varnam) %>%
  summarise(mean = mean(value)) %>%
  arrange(desc(mean))
resamps_tbl$varnam <- factor(resamps_tbl$varnam, levels = rev(tmp$varnam))

gg_training <- resamps_tbl %>%
  ggplot(aes(x = varnam, y = value)) +
  geom_hline(yintercept = 0.5, colour = "darkgrey") +
  geom_boxplot(outlier.shape = 1) +  # c.f. bwplot(resamps, metric = "ROC")
  # stat_summary(fun.data = "mean_cl_normal") +  # c.f. dotplot(resamps, metric = "ROC")
  labs(
    x = "",
    y = "Area under the ROC curve\n(repeated cross-validation resamples)", 
    title = "Training set") +
  coord_flip(ylim = c(0.25, 1))
```

```{r}
predict_twoclass <- function(object, newdata = NULL) {
  if (is(object, "train")) {
    cbind(
      data.frame(pred = predict(object, newdata, type = "raw")), 
      predict(object, newdata, type = "prob"))
  } else if (is(object, "sbf")) {
    predict(fit, newdata)
  }
}
```

```{r}
pred_df <- predict_twoclass(fit, testing)

testing_df <- pData(casecontstudy)[rownames(testing), ]
testing_df$model_pred_class_ <- pred_df$pred
testing_df$model_pred_prob <- pred_df$case
testing_df$model_pred_score <- gtools::logit(testing_df$model_pred_prob)
```

```{r}
require(pROC)
roc_HER2_amplicon_metagene <- roc(casecontcd ~ HER2_amplicon_metagene,
  data = testing_df, levels = c("control", "case"))
roc_model_pred <- roc(casecontcd ~ model_pred_prob,
  data = testing_df, levels = c("control", "case"))
```

```{r}
gg_test <- tibble(
  varnam = c("HER2 amplicon metagene", selected_modelnam), 
  value = c(
    as.numeric(roc_HER2_amplicon_metagene$auc), 
    as.numeric(roc_model_pred$auc))
  ) %>%
  ggplot(aes(x = varnam, y = value)) +
  geom_hline(yintercept = 0.5, colour = "darkgrey") +
  geom_point() +
  labs(x = "", y = "Area under the ROC curve", 
    title = "Test set") +
  coord_flip(ylim = c(0.25, 1))
```

```{r, fig.width = 8, fig.height = 6, fig.cap = "Figure. Performance in the training and test tests as assessed by area under the ROC curve."}
cowplot::plot_grid(gg_training, gg_test, 
  ncol = 1, rel_heights = c(2, 1), align = "v")
```

```{r, fig.width = 12, fig.height = 6, fig.cap = "Figure. ROC curves for the test set."}
layout(matrix(1:2, nrow = 1, byrow = TRUE))

par(bty = "n")
tmp <- plot(roc_HER2_amplicon_metagene, legacy.axes = TRUE,
  col = colNEJM["blue"], lty = "solid", lwd = 1, 
  main = "Test set")
tmp <- plot(roc_model_pred, legacy.axes = TRUE,
  col = colNEJM["red"], lty = "solid", lwd = 1, add = TRUE)
legend("bottomright",
  legend = c(selected_modelnam, "HER2 amplicon metagene"),
  col = colNEJM[c("red", "blue")],
  lty = c("solid", "solid"),
  lwd = 1, bty = "n")
```

```{r}
uni_models <- list()  # initalise

uni_models[["model_pred_class_"]] <- clogit(
  casecontstat ~ model_pred_class_ + strata(setnr),
  data = testing_df)

uni_models[["h2stat"]] <- clogit(
  casecontstat ~ h2stat + strata(setnr),
  data = testing_df)

uni_models[["trtgrp:scale(PAM50PROLIF)"]] <- clogit(
  casecontstat ~ trtgrp:scale(PAM50PROLIF) + strata(setnr),
  data = testing_df)
```

```{r}
model_multi <- clogit(
  casecontstat ~ model_pred_class_ + 
    h2stat + trtgrp:scale(PAM50PROLIF) + strata(setnr),
  data = testing_df)

uni_models %>%
  purrr::map(compile_res_tbl, suffix = "univariate models") %>%
  purrr::reduce(rbind) %>%
  full_join(compile_res_tbl(model_multi, suffix = "multivariable model"), by = "coef") %>%
  knitr::kable(
    row.names = FALSE, digits = 2, 
    caption = paste("Table.",
      "Illustrative statistical inference of the test set.", 
      "HR: hazard ratio; ET, endocrine therapy; CT, chemotherapy; PAM50PROLIF: PAM50 proliferation index.",
      "Statistical significance codes:  0 '\\*\\*\\*' 0.001 '\\*\\*' 0.01 '\\*' 0.05 '.' 0.1 '\\  ' 1."))
```

## Performance in QC substudy

```{r}
esetqc <- qcsubstudy[
  fData(casecontstudy)$probeid, 
  qcsubstudy_primary_endpoint_inclusion$geo_accession]

featureNames(esetqc) <- make.names(fData(esetqc)$probeid)
stopifnot(validObject(esetqc))

qcsubstudy_df <- cbind(
  pData(esetqc)[, "casecontcd", drop = FALSE],
  as.data.frame(t(exprs(esetqc))))

pred_df <- predict_twoclass(fit, qcsubstudy_df)

esetqc$model_pred_class_ <- pred_df$pred
esetqc$model_pred_prob <- pred_df$case
esetqc$model_pred_score <- gtools::logit(esetqc$model_pred_prob)
```

```{r, fig.width = 12, fig.height = 6, fig.cap = "Figure. ROC curves for the QC substudy."}
require(pROC)

layout(matrix(1:2, nrow = 1, byrow = TRUE))
par(bty = "n")

## Original RNA
roc_HER2_amplicon_metagene <- roc(casecontcd ~ HER2_amplicon_metagene,
  data = subset(pData(esetqc), rna_extract == "original"), 
  levels = c("control", "case"))
roc_model_pred <- roc(casecontcd ~ model_pred_score,
  data = subset(pData(esetqc), rna_extract == "original"), 
  levels = c("control", "case"))

tmp <- plot(roc_HER2_amplicon_metagene, legacy.axes = TRUE,
  col = colNEJM["blue"], lty = "solid", lwd = 1, 
  main = "Original RNA")
tmp <- plot(roc_model_pred, legacy.axes = TRUE,
  col = colNEJM["red"], lty = "solid", lwd = 1, add = TRUE)
legend("bottomright",
  legend = c(selected_modelnam, "HER2 amplicon metagene"),
  col = colNEJM[c("red", "blue")],
  lty = c("solid", "solid"),
  lwd = 1, bty = "n")

## Reextracted RNA
roc_HER2_amplicon_metagene <- roc(casecontcd ~ HER2_amplicon_metagene,
  data = subset(pData(esetqc), rna_extract == "reextract"), 
  levels = c("control", "case"))
roc_model_pred <- roc(casecontcd ~ model_pred_score,
  data = subset(pData(esetqc), rna_extract == "reextract"), 
  levels = c("control", "case"))

tmp <- plot(roc_HER2_amplicon_metagene, legacy.axes = TRUE,
  col = colNEJM["blue"], lty = "solid", lwd = 1, 
  main = "Reextracted RNA")
tmp <- plot(roc_model_pred, legacy.axes = TRUE,
  col = colNEJM["red"], lty = "solid", lwd = 1, add = TRUE)
legend("bottomright",
  legend = c(selected_modelnam, "HER2 amplicon metagene"),
  col = colNEJM[c("red", "blue")],
  lty = c("solid", "solid"),
  lwd = 1, bty = "n")
```

```{r derive_setdiffs}
sel_markers <- c(
  "model_pred_score" = selected_modelnam, 
  "HER2_amplicon_metagene" = "HER2 amplicon metagene")

setdiff_tbl <- pData(esetqc) %>%
  ## Identify sets with one case and at least one control
  group_by(rna_extract, setnr) %>%
  mutate(
    n_cases_in_set = sum(casecontcd == "case"),
    n_controls_in_set = sum(casecontcd == "control")) %>%
  dplyr::filter(n_cases_in_set == 1 & n_controls_in_set >= 1) %>%
  ## Derive case-control set differences
  dplyr::select(rna_extract, setnr, casecontcd, one_of(names(sel_markers))) %>%
  gather(markercd, value, one_of(names(sel_markers))) %>%
  mutate(
    markercd = factor(markercd, 
      levels = names(sel_markers)),
    marker = factor(sel_markers[as.character(markercd)], 
      levels = sel_markers)
    ) %>%
  group_by(rna_extract, setnr, markercd, marker, casecontcd) %>%
  summarise(value = mean(value)) %>%
  spread(casecontcd, value) %>%
  mutate(setdiff = case - control)  %>%
  dplyr::select(markercd, marker, setnr, rna_extract, setdiff)
```

```{r, fig.cap = "Figure. Case-control set differences."}
lim0 <- c(-1, 1) * max(abs(setdiff_tbl$setdiff))

colmarkercd <- colNEJM[c("red", "blue")]
names(colmarkercd) <- c("model_pred_score", "HER2_amplicon_metagene")

gg <- setdiff_tbl %>%
  spread(rna_extract, setdiff) %>%
  ggplot(aes(x = original, y = reextract, col = markercd)) +
  geom_vline(xintercept = 0, col = "darkgrey", linetype = "solid") +
  geom_hline(yintercept = 0, col = "darkgrey", linetype = "solid") +
  geom_abline(intercept = 0, slope = 1, col = "darkgrey", linetype = "dashed") +
  geom_point() +
  facet_wrap(~ marker) +
  coord_fixed() +
  scale_colour_manual(values = colmarkercd, guide = FALSE) +
  lims(x = lim0, y = lim0) +
  labs(
    x = "Case-control set difference\nOriginal RNA",
    y = "Case-control set difference\nReextracted RNA")
plot(gg)
```
