---
title: "(draft) Reproduction of biostatistical analyses in Cunha15CanRes"
author: "John Lövrot"
license: "CC BY 4.0"
date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    fig_width: 8
    fig_height: 4
link-citations: yes
bibliography: bibliography.bib
csl: journal-of-the-national-cancer-institute.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = 'graphics/reproduce-cunha15canres-', 
    echo = FALSE, warning = FALSE, message = FALSE)
```

```{r load_project}
setwd("..")
ProjectTemplate::reload.project()
setwd("reports")

select <- dplyr::select  # mask AnnotationDbi::select
```

```{r settings}
ggplot2::theme_set(theme_classic() +
    theme(axis.line.x = element_blank()) + 
    theme(axis.line.y = element_blank()))

coltrtgrp <- colJCO[c("blue", "yellow", "red")]
names(coltrtgrp) <- c("ET", "CT+ET", "CT")

options(knitr.kable.NA = "")
```

```{r}
if (interactive()) {
  ## cf. Tab1
  tab1_symbols <- c("ACVRL1", "PECAM1", "CDH5", "CD34",
    "GDF2", "BMP10", "SMAD6", "TGFBR1", "TGFB1", "TGFB1", "TGFB3")
  
  tibble(
    symbol = tab1_symbols, 
    gse48091 = tab1_symbols %in% fData(gse48091)$symbol,
    gse81954 = tab1_symbols %in% fData(gse81954)$symbol) %>%
    knitr::kable()

  ## Preprocessed data for all probesets are not availabled in GSE48091. 
  ## A non-specific variance filter has been applied, see, 
  ## for example, https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1167774
  ## "A variance filter was employed and probes with inter-quartile-range less 
  ## than 0.5 were excluded, resulting in 28449 retained probes."
  
  ## An in-house full ExpressionSet was used in the analysis for the original article.
}
```

```{r additional_munging}
sel_symbols <- c("ACVRL1", "PECAM1", "CDH5", "CD34", "ESR1")
eset <- casecontstudy[fData(casecontstudy)$symbol %in% sel_symbols, ] %>%
  genefilter::featureFilter()
featureNames(eset) <- as.character(fData(eset)$symbol)
pData(eset) <- cbind(pData(eset), t(exprs(eset)))

pData(casecontstudy) <- pData(eset) %>%
  as_tibble() %>%
  mutate(
    endothelial_metagene = (PECAM1 + CDH5 + CD34) / 3,
    ACVRL1_endothelial_index = ACVRL1 - endothelial_metagene, 
    h2stat = factor(subtypecd %in% "H2", 
      levels = c(FALSE, TRUE), labels = c("other", "H2"))
    ) %>%
  (function(x) data.frame(x, row.names = x$subjid))

stopifnot(validObject(casecontstudy))
```

```{r}
## Derive excess distant metastases (martingale residuals in an empty model)
## Same as clogit, but not method="exact"; now defaults to "efron"
model <- coxph(
  Surv(rep(1, ncol(casecontstudy)), casecontstat) ~ 0 + strata(setnr),
  data =  pData(casecontstudy))
casecontstudy$excessdm <- residuals(model, method = "martingale")
```

# Aim

The aim of this report is to reproduce the statistical analyses of the nested case-control study [GSE48091](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE48091) as part of the original research article by Cunha, Bocci, Lövrot, et al. [@Cunha2015;@Lovrot2016]

Since not all clinical-pathological data is public, in particular disease stage (tumour size and lymph node status), the analyses are not identical to the ones in the article. 

# Background

To be added.

# Data analysis

## Co-expression analysis

Endothelial genes co-expressed with ACVRL1.

```{r}
endothelial_genes <- c("TIE1", "PECAM1", "CD34", "ESAM", "CDH5", "VWF", "FLI1")

eset <- gse48091[fData(gse48091)$symbol %in% c("ACVRL1", endothelial_genes), ] %>%
  genefilter::featureFilter()
featureNames(eset) <- as.character(fData(eset)$symbol)

endothelial_genes %>%
  purrr::map(
    function(symbol) {
    fit <- cor.test(
      x = exprs(eset)[symbol, ], y = exprs(eset)["ACVRL1", ], 
      method = "pearson")
    tibble(
      "Gene name" = symbol,
      "Correlation coefficient" = fit$estimate, 
      "95% CI lower limit" = fit$conf.int[1],
      "95% CI uppper limit" = fit$conf.int[2])
  }) %>%
  purrr::reduce(rbind) %>%
  knitr::kable(
    row.names = FALSE, digits = 2, 
    caption = paste(
      "Table (cf. Suppl. Tab. 1 in the original article).", 
      "Endothelial genes co-expressed with ACVRL1 in the nested case-control study."))
```

## Statistical inference

Univariate and multivariable conditional logistic regression models are used to compare patients developing metastatic disease with patient free from disseminating disase in the nested case-control study, 
where the controls are randomly matched to cases by age, adjuvant systemic therapy and calendar period at diagnosis.

Since the clinical-pathological data on tumour size, lymph node status and HER2-status is not public, the analyses are not identical to the ones in the article. 
HER2-enriched intrinsic subtype as determined by PAM50[@Parker2009] (versus rest) is used as a surrogate for clinical HER2-status.

Moreover, within-therapy-group associations for proliferation as represented by the PAM50-PROLIF index[@Nielsen2010] is added as proliferation is a strong prognostic factor in ER-positive breast cancer (enriched in the endocrine therapy groups) and a (potential) predictive factor for response to chemotherapy, with opposite direction of association with outcome. 
See also illustrative figure in the Supportive information below.

```{r}
uni_models <- list()  # initalise

uni_models[["scale(ACVRL1)"]] <- clogit(
  casecontstat ~ scale(ACVRL1) + strata(setnr),
  data = pData(casecontstudy))

uni_models[["scale(endothelial_metagene)"]] <- clogit(
  casecontstat ~ scale(endothelial_metagene) + strata(setnr),
  data = pData(casecontstudy))

uni_models[["scale(ACVRL1_endothelial_index)"]] <- clogit(
  casecontstat ~ scale(ACVRL1_endothelial_index) + strata(setnr),
  data = pData(casecontstudy))

uni_models[["h2stat"]] <- clogit(
  casecontstat ~ h2stat + strata(setnr),
  data = pData(casecontstudy))

uni_models[["trtgrp:scale(PAM50PROLIF)"]] <- clogit(
  casecontstat ~ trtgrp:scale(PAM50PROLIF) + strata(setnr),
  data = pData(casecontstudy))
```

```{r}
model_multiA <- clogit(
  casecontstat ~ scale(ACVRL1) + scale(endothelial_metagene) + 
    h2stat + trtgrp:scale(PAM50PROLIF) + strata(setnr),
  data = pData(casecontstudy))

model_multiB <- clogit(
  casecontstat ~ scale(ACVRL1_endothelial_index) + 
    h2stat + trtgrp:scale(PAM50PROLIF) + strata(setnr),
  data = pData(casecontstudy))

uni_models %>%
  purrr::map(compile_res_tbl, suffix = "univariate models") %>%
  purrr::reduce(rbind) %>%
  full_join(compile_res_tbl(model_multiA, suffix = "multivariable model A"), by = "coef") %>%
  full_join(compile_res_tbl(model_multiB, suffix = "multivariable model B"), by = "coef") %>%
  knitr::kable(
    row.names = FALSE, digits = 2, 
    caption = paste("Table (cf. Tab. 1 in the original article).",
      "Univariate and multivariable conditional logistic regression models of the nested case-control study.", 
      "Per standard-deviation hazard ratios (HR) for continuous variables.", 
      "ET, endocrine therapy; CT, chemotherapy; H2, HER2-enriched intrinsic subtype; PAM50PROLIF: PAM50 proliferation index.",
      "Statistical significance codes:  0 '\\*\\*\\*' 0.001 '\\*\\*' 0.01 '\\*' 0.05 '.' 0.1 '\\  ' 1."))
    
```

Endothelial metagene expression is the average expression of the prototypical endothelial cell markers PECAM1, CDH5, and CD34.

# Supportive information

Relating ACVRL1 to prototypical endothelial gene-expression makes the association with risk of disseminating disease coherent across treatment groups.

```{r, fig.height = 8, fig.cap = "Figure (cf. Suppl. Fig. 3A and 3B in the original article). (A) Exploratory plots of excess distant metastases versus the ACVRL1:endothelial metagene index, stratified by adjuvant systemic therapy. ET, endocrine therapy; CT, chemotherapy. (B) Box plots show similar manifestation of the ACVRL1:endothelial metagene across all molecular subtypes of breast cancer. LA, luminal A; LB, luminal B; H2, HER2-enriched; BL, basal-like; NBL, normal breast-like."}
gg1 <- pData(casecontstudy) %>%
  ggplot(aes(x = scale(ACVRL1_endothelial_index), y = excessdm)) +
  geom_point(shape = 1) +
  geom_smooth(aes(col = trtgrp), method = "loess") +
  facet_grid(. ~ trtgrp) +
  scale_colour_manual(values = coltrtgrp) +
  labs(
    x = "ACVRL1:endothelial metagene index\n(centered and scaled)",
    y = "Excess distant metastases", col = "Adjuvant systemic therapy") +
  theme(legend.position = "bottom")

gg2 <- pData(casecontstudy) %>%
  ggplot(aes(x = subtypecd, y = scale(ACVRL1_endothelial_index))) +
  geom_boxplot(aes(fill = subtypecd), outlier.shape = 1, notch = TRUE, varwidth = TRUE) +
  scale_fill_manual(values = colsubtypecd, guide = FALSE) +
  labs(
    x = "Intrinsic molecular subtype",
    y = "ACVRL1:endothelial metagene index\n(centered and scaled)")

cowplot::ggdraw() +
  cowplot::draw_plot(gg1, x = 0, y = 0.5, w = 1, h = 0.5) +
  cowplot::draw_plot(gg2, x = 0, y = 0, w = 0.5, h = 0.5) +
  cowplot::draw_plot_label(c("A", "B"), x = c(0, 0), y = c(1, 0.5))
```

The shape of the smoother in an excess distant metastases plot indicates the form of an association between the gene expression variable and risk of metastatic disease. 
Mathematically, the excess distant metastases are martingale residuals in a null conditional logistic regression model.

```{r, fig.height = 6, fig.cap = "Figure. Illustration of the association with outcome for proliferation status of the tumour as examplified by the PAM50-PROLIF index."}
gg1 <- pData(casecontstudy) %>%
  ggplot(aes(x = PAM50PROLIF, y = excessdm)) + 
  geom_point(shape = 1) +
  geom_smooth(method = "loess", col = "darkred") +
  labs(y = "Excess distant metastases")

gg2 <- pData(casecontstudy) %>%
  ggplot(aes(x = trtgrp, y = PAM50PROLIF)) + 
  geom_boxplot(outlier.shape = 1, varwidth = TRUE) +
  labs(x = "") +
  coord_flip()

cowplot::plot_grid(
  gg1, 
  gg1 + facet_wrap(~ trtgrp), 
  gg2,
  labels = "AUTO", ncol = 2, rel_widths = c(1, 2))
```

# Addendum

Addendum/Correction[@Lovrot2016], 

```{r}
sel_symbols <- c("ACVRL1", "PECAM1", "CDH5", "CD34", "ESR1", "ERBB2")

eset <- qcsubstudy[fData(qcsubstudy)$symbol %in% sel_symbols, ] %>%
  averageExprsByFDataVar("symbol")
featureNames(eset) <- as.character(fData(eset)$symbol)

pData(qcsubstudy) <- cbind(pData(qcsubstudy), t(exprs(eset)))

stopifnot(validObject(qcsubstudy))
```

> The correlation between gene expression data for original and reextracted RNA is excellent for key breast cancer genes, for example, ESR1 (r = 0.95) and ERBB2 (r = 0.96).

```{r}
sel_symbols <- c("ESR1", "ERBB2")

pData(qcsubstudy) %>%
  select(tumorid, rna_extract, one_of(sel_symbols)) %>%
  gather(symbol, value, one_of(sel_symbols)) %>%
  mutate(symbol = factor(symbol, sel_symbols)) %>%
  spread(rna_extract, value) %>%
  group_by(symbol) %>%
  summarise(r = cor(original, reextract, method = "pearson")) %>%
  rename(c(
    symbol = "Gene name",
    r = "Correlation coefficient")) %>%
  knitr::kable(digits = 3)
```

> Bridging the primary comparison, case–control set differences (n = 40) for ACVRL1 and the ACVRL1:endothelial metagene index that we reported are consistent between the two extractions (Fig. 1). 
> A case–control set difference is the value for the case minus the (average) value of the matched control(s).

```{r}
pData(qcsubstudy) <- pData(qcsubstudy) %>%
  as_tibble() %>%
  mutate(
    endothelial_metagene = PECAM1 + CDH5 + CD34,  # unfortunately, this incorrect derivation was used for the addendum
    # endothelial_metagene = (PECAM1 + CDH5 + CD34) / 3,  # correct derivation; reassuringly, the consistency remains.
    ACVRL1_endothelial_index = ACVRL1 - endothelial_metagene
  ) %>%
  (function(x) data.frame(x, row.names = x$geo_accession))

stopifnot(validObject(qcsubstudy))
```

```{r}
sel_markers <- c("ACVRL1", "ACVRL1_endothelial_index")

setdiff_tbl <- pData(qcsubstudy) %>%
  ## Keep sets with one case and at least one control
  filter(complete_casecontset == TRUE) %>%
  ## Derive case-control set differences
  select(rna_extract, setnr, casecontcd, one_of(sel_markers)) %>%
  gather(marker, value, one_of(sel_markers)) %>%
  group_by(rna_extract, setnr, marker, casecontcd) %>%
  summarise(value = mean(value)) %>%
  spread(casecontcd, value) %>%
  mutate(setdiff = case - control) %>%
  ## Spread into columns for original and re-extracted RNA
  select(marker, setnr, rna_extract, setdiff) %>%
  spread(rna_extract, setdiff)
```

```{r setdiffs, fig.height=4, fig.width=8, fig.cap = "Figure. Case-control set differences."}
ggplot_setdiff <- function(data) {
  lim0 <- c(-1, 1) * max(abs(c(data$original, data$reextract)))
  ggplot(data, aes(x = original, y = reextract)) +
    geom_vline(xintercept = 0, col = "darkgrey", linetype = "solid") +
    geom_hline(yintercept = 0, col = "darkgrey", linetype = "solid") +
    geom_abline(intercept = 0, slope = 1, col = "darkgrey", linetype = "dashed") +
    geom_point() +
    facet_wrap(~ marker) +
    coord_fixed() +
    lims(x = lim0, y = lim0) +
    labs(
      x = "Case-control set difference\noriginal RNA",
      y = "Case-control set difference\nreextracted RNA")
}

cowplot::plot_grid(
  ggplot_setdiff(filter(setdiff_tbl, marker == "ACVRL1")),
  ggplot_setdiff(filter(setdiff_tbl, marker == "ACVRL1_endothelial_index")),
  labels = "AUTO", ncol = 2)
```

# Additional information

Addtional illustrative predictive models.

```{r}
library(caret)
library(glmnet)
library(randomForest)
library(pROC)
```

```{r}
trellis.par.set(caretTheme())
```

```{r}
load(file.path("..", "cache", "caret", "training.RData"))
```

## Training of illustrative models

The full study is split into a training set (2/3) and a test set (1/3), 
with the QC substudy as part of the test set.

```{r, echo = FALSE}
casecontsets <- unique(casecontstudy$setnr)
tibble(
  casecontset = casecontsets, 
  partition = factor(
    casecontset %in% pData(casecontstudy)[rownames(training), "setnr"], 
    levels = c(TRUE, FALSE), 
    labels = c("Training set", "Test set")), 
  in_qcsubstudy = factor(
    casecontset %in% subset(pData(qcsubstudy), complete_casecontset)$setnr, 
    levels = c(TRUE, FALSE), 
    labels = c("QC substudy", "(rest)"))) %>%
  dplyr::select(in_qcsubstudy, partition) %>%
  table() %>%
  addmargins() %>%
  knitr::kable(caption = "Table. Number of case-control sets in each partition of the full study.")
```

A panel of models are trained and tuned based on the training set using repeated cross-validation, a resampling technique.
For details, see [reproduce-cunha15canres/addendum-train-illustrative-models.html](./addendum-train-illustrative-models.html)

## Performance in training and test sets

```{r}
resamps <- resamples(fit_list)
```

```{r}
## Select model based on comparison of performance in training set
if (interactive()) 
  dotplot(resamps, metric = "ROC")

selected_modelnam <- "Logistic Regression with Elastic Net Regularisation"
fit <- fit_list[[selected_modelnam]]
```

Selected model: 
```{r, results = "asis"}
cat(selected_modelnam)
```

```{r}
resamps_tbl <- resamps %>%
  as.data.frame() %>%
  tidyr::gather(varnam, value, -Resample) %>%
  tidyr::separate(Resample, c("fold", "rep"))

## Find ordering of models
tmp <- resamps_tbl %>% 
  group_by(varnam) %>%
  summarise(mean = mean(value)) %>%
  arrange(desc(mean))
resamps_tbl$varnam <- factor(resamps_tbl$varnam, levels = rev(tmp$varnam))

gg_training <- resamps_tbl %>%
  ggplot(aes(x = varnam, y = value)) +
  geom_hline(yintercept = 0.5, colour = "darkgrey") +
  geom_boxplot(outlier.shape = 1) +  # c.f. bwplot(resamps, metric = "ROC")
  # stat_summary(fun.data = "mean_cl_normal") +  # c.f. dotplot(resamps, metric = "ROC")
  labs(
    x = "",
    y = "Area under the ROC curve\n(repeated cross-validation resamples)", 
    title = "Training set") +
  coord_flip(ylim = c(0.25, 1))
```

```{r}
predict_twoclass <- function(object, newdata = NULL) {
  if (is(object, "train")) {
    cbind(
      data.frame(pred = predict(object, newdata, type = "raw")), 
      predict(object, newdata, type = "prob"))
  } else if (is(object, "sbf")) {
    predict(fit, newdata)
  }
}
```

```{r}
pred_df <- predict_twoclass(fit, testing)

testing_df <- pData(casecontstudy)[rownames(testing), ]
testing_df$model_pred_class_ <- pred_df$pred
testing_df$model_pred_prob <- pred_df$case
testing_df$model_pred_score <- gtools::logit(testing_df$model_pred_prob)
```

```{r}
require(pROC)
roc_HER2_amplicon_metagene <- roc(casecontcd ~ HER2_amplicon_metagene,
  data = testing_df, levels = c("control", "case"))
roc_model_pred <- roc(casecontcd ~ model_pred_prob,
  data = testing_df, levels = c("control", "case"))
```

```{r}
gg_test <- tibble(
  varnam = c("HER2 amplicon metagene", selected_modelnam), 
  value = c(
    as.numeric(roc_HER2_amplicon_metagene$auc), 
    as.numeric(roc_model_pred$auc))
  ) %>%
  ggplot(aes(x = varnam, y = value)) +
  geom_hline(yintercept = 0.5, colour = "darkgrey") +
  geom_point() +
  labs(x = "", y = "Area under the ROC curve", 
    title = "Test set") +
  coord_flip(ylim = c(0.25, 1))
```

```{r, fig.width = 8, fig.height = 6, fig.cap = "Figure. Performance in the training and test tests as assessed by area under the ROC curve."}
cowplot::plot_grid(gg_training, gg_test, 
  ncol = 1, rel_heights = c(2, 1), align = "v")
```

```{r, fig.width = 12, fig.height = 6, fig.cap = "Figure. ROC curves for the test set."}
layout(matrix(1:2, nrow = 1, byrow = TRUE))

par(bty = "n")
tmp <- plot(roc_HER2_amplicon_metagene, legacy.axes = TRUE,
  col = colNEJM["blue"], lty = "solid", lwd = 1, 
  main = "Test set")
tmp <- plot(roc_model_pred, legacy.axes = TRUE,
  col = colNEJM["red"], lty = "solid", lwd = 1, add = TRUE)
legend("bottomright",
  legend = c(selected_modelnam, "HER2 amplicon metagene"),
  col = colNEJM[c("red", "blue")],
  lty = c("solid", "solid"),
  lwd = 1, bty = "n")
```

```{r}
uni_models <- list()  # initalise

uni_models[["model_pred_class_"]] <- clogit(
  casecontstat ~ model_pred_class_ + strata(setnr),
  data = testing_df)

uni_models[["h2stat"]] <- clogit(
  casecontstat ~ h2stat + strata(setnr),
  data = testing_df)

uni_models[["trtgrp:scale(PAM50PROLIF)"]] <- clogit(
  casecontstat ~ trtgrp:scale(PAM50PROLIF) + strata(setnr),
  data = testing_df)
```

```{r}
model_multi <- clogit(
  casecontstat ~ model_pred_class_ + 
    h2stat + trtgrp:scale(PAM50PROLIF) + strata(setnr),
  data = testing_df)

uni_models %>%
  purrr::map(compile_res_tbl, suffix = "univariate models") %>%
  purrr::reduce(rbind) %>%
  full_join(compile_res_tbl(model_multi, suffix = "multivariable model"), by = "coef") %>%
  knitr::kable(
    row.names = FALSE, digits = 2, 
    caption = paste("Table.",
      "Illustrative statistical inference of the test set.", 
      "HR: hazard ratio; ET, endocrine therapy; CT, chemotherapy; PAM50PROLIF: PAM50 proliferation index.",
      "Statistical significance codes:  0 '\\*\\*\\*' 0.001 '\\*\\*' 0.01 '\\*' 0.05 '.' 0.1 '\\  ' 1."))
```

## Performance in QC substudy

```{r}
esetqc <- qcsubstudy[
  fData(casecontstudy)$probeid, 
  qcsubstudy$complete_casecontset]

featureNames(esetqc) <- make.names(fData(esetqc)$probeid)
stopifnot(validObject(esetqc))

qcsubstudy_df <- cbind(
  pData(esetqc)[, "casecontcd", drop = FALSE],
  as.data.frame(t(exprs(esetqc))))

pred_df <- predict_twoclass(fit, qcsubstudy_df)

esetqc$model_pred_class_ <- pred_df$pred
esetqc$model_pred_prob <- pred_df$case
esetqc$model_pred_score <- gtools::logit(esetqc$model_pred_prob)
```

```{r, fig.width = 12, fig.height = 6, fig.cap = "Figure. ROC curves for the QC substudy."}
require(pROC)

layout(matrix(1:2, nrow = 1, byrow = TRUE))
par(bty = "n")

## Original RNA
roc_HER2_amplicon_metagene <- roc(casecontcd ~ HER2_amplicon_metagene,
  data = subset(pData(esetqc), rna_extract == "original"), 
  levels = c("control", "case"))
roc_model_pred <- roc(casecontcd ~ model_pred_score,
  data = subset(pData(esetqc), rna_extract == "original"), 
  levels = c("control", "case"))

tmp <- plot(roc_HER2_amplicon_metagene, legacy.axes = TRUE,
  col = colNEJM["blue"], lty = "solid", lwd = 1, 
  main = "Original RNA")
tmp <- plot(roc_model_pred, legacy.axes = TRUE,
  col = colNEJM["red"], lty = "solid", lwd = 1, add = TRUE)
legend("bottomright",
  legend = c(selected_modelnam, "HER2 amplicon metagene"),
  col = colNEJM[c("red", "blue")],
  lty = c("solid", "solid"),
  lwd = 1, bty = "n")

## Reextracted RNA
roc_HER2_amplicon_metagene <- roc(casecontcd ~ HER2_amplicon_metagene,
  data = subset(pData(esetqc), rna_extract == "reextract"), 
  levels = c("control", "case"))
roc_model_pred <- roc(casecontcd ~ model_pred_score,
  data = subset(pData(esetqc), rna_extract == "reextract"), 
  levels = c("control", "case"))

tmp <- plot(roc_HER2_amplicon_metagene, legacy.axes = TRUE,
  col = colNEJM["blue"], lty = "solid", lwd = 1, 
  main = "Reextracted RNA")
tmp <- plot(roc_model_pred, legacy.axes = TRUE,
  col = colNEJM["red"], lty = "solid", lwd = 1, add = TRUE)
legend("bottomright",
  legend = c(selected_modelnam, "HER2 amplicon metagene"),
  col = colNEJM[c("red", "blue")],
  lty = c("solid", "solid"),
  lwd = 1, bty = "n")
```

```{r derive_setdiffs}
sel_markers <- c(
  "model_pred_score" = selected_modelnam, 
  "HER2_amplicon_metagene" = "HER2 amplicon metagene")

setdiff_tbl <- pData(esetqc) %>%
  ## Identify sets with one case and at least one control
  group_by(rna_extract, setnr) %>%
  mutate(
    n_cases_in_set = sum(casecontcd == "case"),
    n_controls_in_set = sum(casecontcd == "control")) %>%
  dplyr::filter(n_cases_in_set == 1 & n_controls_in_set >= 1) %>%
  ## Derive case-control set differences
  dplyr::select(rna_extract, setnr, casecontcd, one_of(names(sel_markers))) %>%
  gather(markercd, value, one_of(names(sel_markers))) %>%
  mutate(
    markercd = factor(markercd, 
      levels = names(sel_markers)),
    marker = factor(sel_markers[as.character(markercd)], 
      levels = sel_markers)
    ) %>%
  group_by(rna_extract, setnr, markercd, marker, casecontcd) %>%
  summarise(value = mean(value)) %>%
  spread(casecontcd, value) %>%
  mutate(setdiff = case - control)  %>%
  dplyr::select(markercd, marker, setnr, rna_extract, setdiff)
```

```{r, fig.cap = "Figure. Case-control set differences."}
lim0 <- c(-1, 1) * max(abs(setdiff_tbl$setdiff))

colmarkercd <- colNEJM[c("red", "blue")]
names(colmarkercd) <- c("model_pred_score", "HER2_amplicon_metagene")

gg <- setdiff_tbl %>%
  spread(rna_extract, setdiff) %>%
  ggplot(aes(x = original, y = reextract, col = markercd)) +
  geom_vline(xintercept = 0, col = "darkgrey", linetype = "solid") +
  geom_hline(yintercept = 0, col = "darkgrey", linetype = "solid") +
  geom_abline(intercept = 0, slope = 1, col = "darkgrey", linetype = "dashed") +
  geom_point() +
  facet_wrap(~ marker) +
  coord_fixed() +
  scale_colour_manual(values = colmarkercd, guide = FALSE) +
  lims(x = lim0, y = lim0) +
  labs(
    x = "Case-control set difference\nOriginal RNA",
    y = "Case-control set difference\nReextracted RNA")
plot(gg)
```

# R session information

```{r}
print(sessionInfo(), locale = FALSE)
```

- - -

&copy; 2017 John Lövrot.  
This work is licensed under a [Creative Commons Attribution 4.0 International License](http://creativecommons.org/licenses/by/4.0/).  
The source code is available at [github.com/lovrot/reproduce-cunha15canres](http://github.com/lovrot/reproduce-cunha15canres).  
Version `r format(read.dcf("../description.dcf")[1, "version"])`

# References
